---
title: "Using synthpop to generate synthetic data for agentic coding assistants"
author: "Billy Quilty"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Using synthpop to generate synthetic data for agentic coding assistants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 10,
  fig.height = 8,
  dpi = 100
)

# Consistent colour scheme throughout
col_real <- "#2166ac"
col_synth <- "#b2182b"
```

## The problem

Research analysis projects that handle patient-identifiable data face a
practical tension when using AI coding agents such as Claude Code, GitHub
Copilot, or Cursor. The agent needs to see data to develop the pipeline, but
doing so sends that data to a remote API. Deny rules in the agent's
configuration can block direct file reads, but they are not watertight: shell
output, error messages, and script results can still leak real values into the
conversation context.

Synthetic data offers a cleaner boundary. Generate a version of each dataset
that preserves the schema and statistical properties of the original, develop
the pipeline against that, and run the final analysis on real data outside the
agent. The agent never sees a real record.

This vignette walks through `synthpop`'s functions for this workflow: synthesis,
diagnostics, disclosure control, and integration into an analysis project.

## Synthesising a dataset

`syn()` is the core function. It fits a model to each column of the real data
(conditional on the columns that precede it in the visit sequence) and draws
new records from those models.

```{r}
library(synthpop)

syn_mtcars <- syn(mtcars, method = "parametric", seed = 2024)
```

The result is a `synds` object. The synthetic data frame lives in `$syn`:

```{r}
identical(dim(syn_mtcars$syn), dim(mtcars))
identical(names(syn_mtcars$syn), names(mtcars))
```

### Preparing categorical variables

`synthpop` warns when numeric columns have 5 or fewer unique values, because
these are likely categorical. Converting them to factors before synthesis
ensures they are modelled correctly (logistic/multinomial rather than linear)
and that the synthetic values stay within the observed levels.

Use `numtocat.syn()` to convert these automatically, or do it manually:

```{r}
mtcars_prep <- mtcars
mtcars_prep$cyl <- factor(mtcars_prep$cyl)
mtcars_prep$vs <- factor(mtcars_prep$vs)
mtcars_prep$am <- factor(mtcars_prep$am)
mtcars_prep$gear <- factor(mtcars_prep$gear)

syn_mtcars_f <- syn(mtcars_prep, method = "parametric", seed = 2024)
```

The synthetic data now has the correct factor levels:

```{r}
table(syn_mtcars_f$syn$cyl)
table(mtcars_prep$cyl)
```

## Choosing a method

`synthpop` offers two broad approaches, set with the `method` argument:

- **`"parametric"`**: fits linear models (for numeric columns) and logistic
  regression (for factors). Preserves pairwise correlations well. Faster.
- **`"cart"`** (the default): fits classification and regression trees. Better
  for datasets with complex interactions, non-linear relationships, or strong
  subgroup structure (e.g. disease subtypes with different correlation
  patterns).

You can also set the method per column, mixing approaches where appropriate.

```{r}
syn_param <- syn(mtcars_prep, method = "parametric", seed = 2024)
syn_cart <- syn(mtcars_prep, method = "cart", seed = 2024)
```

## Comparing synthetic and real data

### Pairwise comparison

A `GGally::ggpairs` plot with real and synthetic data overlaid gives an
immediate visual summary of how well all pairwise relationships are preserved.
The upper triangle shows per-group correlations, the diagonal shows density
overlays, and the lower triangle shows scatterplots.

```{r fig.width = 12, fig.height = 10}
if (requireNamespace("GGally", quietly = TRUE)) {
  cols <- c("mpg", "cyl", "disp", "hp", "wt", "am")
  real_df <- mtcars_prep[, cols]
  syn_df <- syn_param$syn[, cols]
  real_df$Source <- "Real"
  syn_df$Source <- "Synthetic"
  combined <- rbind(real_df, syn_df)
  combined$Source <- factor(combined$Source, levels = c("Real", "Synthetic"))

  GGally::ggpairs(
    combined,
    columns = seq_along(cols),
    mapping = ggplot2::aes(colour = Source, alpha = Source),
    upper = list(continuous = "cor"),
    lower = list(
      continuous = GGally::wrap("points", size = 0.8),
      combo = GGally::wrap("box_no_facet")
    ),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.4)),
    title = "Real vs Synthetic: mtcars (parametric, with factor columns)",
    progress = FALSE,
    legend = c(1, 1)
  ) +
    ggplot2::scale_colour_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_fill_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_alpha_manual(values = c(Real = 0.5, Synthetic = 0.5), guide = "none") +
    ggplot2::theme_minimal(base_size = 9)
}
```

For datasets with mixed types (numeric and factor), the plot adapts with
boxplots and bar charts:

```{r fig.width = 12, fig.height = 10}
syn_iris <- syn(iris, method = "cart", seed = 2024)

if (requireNamespace("GGally", quietly = TRUE)) {
  real_ir <- iris
  syn_ir <- syn_iris$syn
  real_ir$Source <- "Real"
  syn_ir$Source <- "Synthetic"
  combined_ir <- rbind(real_ir, syn_ir)
  combined_ir$Source <- factor(combined_ir$Source, levels = c("Real", "Synthetic"))

  GGally::ggpairs(
    combined_ir,
    columns = 1:5,
    mapping = ggplot2::aes(colour = Source, alpha = Source),
    upper = list(continuous = "cor"),
    lower = list(
      continuous = GGally::wrap("points", size = 0.8),
      combo = GGally::wrap("box_no_facet")
    ),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.4)),
    title = "Real vs Synthetic: iris (CART)",
    progress = FALSE,
    legend = c(1, 1)
  ) +
    ggplot2::scale_colour_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_fill_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_alpha_manual(values = c(Real = 0.5, Synthetic = 0.5), guide = "none") +
    ggplot2::theme_minimal(base_size = 9)
}
```

### Per-variable comparison

`synthpop`'s built-in `compare()` produces overlaid density plots for numeric
variables and grouped bar charts for factors. It also reports the propensity
score utility statistic ($S_{pMSE}$) for each variable.

```{r fig.height = 10}
compare(syn_param, mtcars_prep, cols = c(col_real, col_synth))
```

For iris with CART:

```{r fig.height = 8}
compare(syn_iris, iris, cols = c(col_real, col_synth))
```

### Handling missingness

`synthpop` preserves missingness patterns. `airquality` has `NA` values in
`Ozone` and `Solar.R`:

```{r fig.height = 6}
syn_aq <- syn(airquality, method = "parametric", seed = 2024)
compare(syn_aq, airquality, cols = c(col_real, col_synth))
```

```{r}
na_comparison <- data.frame(
  column = names(airquality),
  real_na = colSums(is.na(airquality)),
  synthetic_na = colSums(is.na(syn_aq$syn))
)
na_comparison[na_comparison$real_na > 0, ]
```

### Utility measures

`utility.gen()` trains a classifier to distinguish real from synthetic records.
A pMSE ratio near 1 means the synthetic data is about as distinguishable from
the real data as a random split would be.

```{r}
ug_param <- utility.gen(syn_param, mtcars_prep, print.flag = FALSE)
ug_cart <- utility.gen(syn_cart, mtcars_prep, print.flag = FALSE)

data.frame(
  method = c("Parametric", "CART"),
  pMSE = round(c(ug_param$pMSE, ug_cart$pMSE), 4),
  S_pMSE = round(c(ug_param$S_pMSE, ug_cart$S_pMSE), 2),
  SPECKS = round(c(ug_param$SPECKS, ug_cart$SPECKS), 3)
)
```

`S_pMSE` is the pMSE standardised by its null expectation: values near 1
indicate the synthetic data is about as distinguishable as a random split of
the real data. `SPECKS` is the Kolmogorov-Smirnov statistic on the propensity
scores (0 = indistinguishable, 0.5 = maximally distinguishable).

### Correlation preservation

synthpop does not have a built-in correlation comparison, but it is
straightforward to compute. We use only the numeric columns here:

```{r}
num_cols <- names(mtcars_prep)[sapply(mtcars_prep, is.numeric)]
cor_real <- cor(mtcars_prep[num_cols], use = "complete.obs")
cor_param <- cor(syn_param$syn[num_cols], use = "complete.obs")
cor_cart <- cor(syn_cart$syn[num_cols], use = "complete.obs")

idx <- lower.tri(cor_real)

data.frame(
  method = c("Parametric", "CART"),
  mean_abs_diff = round(c(
    mean(abs(cor_real[idx] - cor_param[idx])),
    mean(abs(cor_real[idx] - cor_cart[idx]))
  ), 3),
  max_abs_diff = round(c(
    max(abs(cor_real[idx] - cor_param[idx])),
    max(abs(cor_real[idx] - cor_cart[idx]))
  ), 3)
)
```

### Comparing model fits

`lm.synds()` fits a linear model to the synthetic data, and `compare()` shows
the coefficients alongside the real-data fit. Overlapping confidence intervals
mean the synthetic data would lead to the same qualitative conclusions.

```{r fig.width = 8, fig.height = 5}
fit_syn <- lm.synds(mpg ~ wt + hp + cyl, data = syn_param)
compare(fit_syn, mtcars_prep)
```

## Disclosure control

### Replicated uniques

A replicated unique is a synthetic record that matches a unique real record on
all columns. These are the most direct re-identification risk.

```{r}
ru_param <- replicated.uniques(syn_param, mtcars_prep)
ru_cart <- replicated.uniques(syn_cart, mtcars_prep)

data.frame(
  method = c("Parametric", "CART"),
  replicated_uniques = c(
    ru_param$res_tab["Replicated uniques", "Number"],
    ru_cart$res_tab["Replicated uniques", "Number"]
  ),
  total_records = nrow(mtcars_prep)
)
```

CART tends to produce more because tree leaves can copy small groups of records.
This is more visible with datasets that have well-separated clusters:

```{r}
ru_iris <- replicated.uniques(syn_iris, iris)

data.frame(
  dataset = "iris (CART)",
  replicated_uniques = ru_iris$res_tab["Replicated uniques", "Number"],
  total_records = nrow(iris),
  pct = round(ru_iris$res_tab["Replicated uniques", "%"], 1)
)
```

### Identity and attribute disclosure

`disclosure()` checks whether knowing a set of quasi-identifier columns (keys)
lets an attacker identify an individual or infer a sensitive attribute:

```{r}
disc <- disclosure(syn_param, mtcars_prep,
                   keys = c("cyl", "hp", "wt"), target = "mpg")
```

The key metrics from `disc$ident` and `disc$attrib`:

```{r}
data.frame(
  measure = c("Identity disclosure (%)", "Attribute disclosure (%)"),
  baseline = round(c(disc$ident[1, 1], disc$attrib[1, 1]), 1),
  synthetic = round(c(disc$ident[2, 1], disc$attrib[2, 1]), 1)
)
```

A good synthesis should show synthetic disclosure well below the baseline.

### Removing high-risk records

If replicated uniques are found, `sdc()` removes or replaces them:

```{r}
syn_iris_sdc <- sdc(syn_iris, iris, rm.replicated.uniques = TRUE)
ru_after <- replicated.uniques(syn_iris_sdc, iris)

data.frame(
  stage = c("Before SDC", "After SDC"),
  replicated_uniques = c(
    ru_iris$res_tab["Replicated uniques", "Number"],
    ru_after$res_tab["Replicated uniques", "Number"]
  )
)
```

## Data governance considerations

Synthetic data is not anonymised data. Before using synthetic data with an
external API:

**Individual-level risk.** Remove direct identifiers (names, addresses, NHS
numbers, dates of birth) before synthesis. Run `replicated.uniques()` and
`disclosure()` on the output. Use `sdc()` to remove any replicated uniques.

**Population-level risk.** Synthetic data preserves the joint distributions and
correlations of the original cohort. For large, common-condition datasets
(thousands of records, widely studied diseases), this is unlikely to be
sensitive. For small or rare-disease cohorts at a named institution, the
statistical structure itself can be sensitive: a synthetic dataset that
preserves the relationship between genotype and outcome for 12 patients is
operationally the same information as the real data for anyone who knows the
cohort exists. In these cases, synthesis does not solve the governance problem.

The practical test: if you would be comfortable sharing a detailed set of
summary statistics and cross-tabulations with the API, the synthetic data is
comparable in information content. If you would not, synthesis does not solve
the problem.

**What your DPO needs to see.** A description of the synthesis method,
confirmation that identifiers were removed, the `replicated.uniques()` and
`disclosure()` output, the data processor's retention and training-use
policies, and a documented risk assessment.

## Integration into an analysis workflow

### Generation script

Run this outside the agent, in a regular R session:

```r
# scripts/00_generate_synthetic.R
library(here)
library(synthpop)
library(readr)

raw_dir <- here("data", "raw")
syn_dir <- here("data", "synthetic")
if (!dir.exists(syn_dir)) dir.create(syn_dir, recursive = TRUE)

for (file in list.files(raw_dir, pattern = "\\.csv$", full.names = TRUE)) {
  real <- read_csv(file, show_col_types = FALSE)
  result <- syn(as.data.frame(real), method = "parametric", seed = 2024)

  # Remove replicated uniques
  ru <- replicated.uniques(result, as.data.frame(real))
  if (ru$res_tab["Replicated uniques", "Number"] > 0) {
    result <- sdc(result, as.data.frame(real), rm.replicated.uniques = TRUE)
  }

  write_csv(result$syn, file.path(syn_dir, basename(file)))
}
```

### Switching between synthetic and real data

Use an environment variable so the safe default is synthetic:

```r
use_synthetic <- toupper(Sys.getenv("SYNTHETIC_DATA", "TRUE")) %in%
  c("TRUE", "1", "YES", "")

data_dir <- if (use_synthetic) {
  here::here("data", "synthetic")
} else {
  here::here("data", "raw")
}
```

To run on real data:

```bash
SYNTHETIC_DATA=false Rscript scripts/00_run_pipeline.R
```

### Agent permissions

Deny reads of real and processed data, allow synthetic:

```json
{
  "permissions": {
    "deny": [
      "Read(./data/raw/**)",
      "Read(./data/processed/**)"
    ]
  }
}
```

The deny rules shown here are for Claude Code's `.claude/settings.json`.
Other agents will have their own configuration mechanisms.

## Summary

| synthpop function | Purpose |
|---|---|
| `syn()` | Generate synthetic data |
| `compare()` | Per-variable distribution comparison |
| `lm.synds()` / `glm.synds()` | Fit models to synthetic data |
| `compare.fit.synds()` | Compare model fits (real vs synthetic) |
| `utility.gen()` | Propensity score utility measure |
| `replicated.uniques()` | Check for records matching unique real individuals |
| `disclosure()` | Identity and attribute disclosure risk |
| `sdc()` | Remove or replace high-risk synthetic records |
