---
title: "Using synthpop to generate synthetic data for agentic coding assistants"
author: "Billy Quilty"
date: today
format:
  html:
    toc: true
    toc-depth: 3
    code-fold: false
    embed-resources: true
    fig-width: 10
    fig-height: 8
    fig-dpi: 150
execute:
  warning: false
  message: false
vignette: >
  %\VignetteIndexEntry{Using synthpop to generate synthetic data for agentic coding assistants}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false

devtools::load_all(here::here())

col_real <- "#6680c0"
col_synth <- "#fc8d62"

theme_synth <- ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(
    plot.title = ggplot2::element_text(size = 13, face = "bold"),
    strip.text = ggplot2::element_text(size = 10, face = "bold"),
    panel.grid.minor = ggplot2::element_blank(),
    panel.grid.major = ggplot2::element_line(
      linewidth = 0.3, colour = "grey90"
    ),
    legend.position = "top",
    legend.title = ggplot2::element_blank()
  )

cor_text <- function(data, mapping, ...) {
  x <- GGally::eval_data_col(data, mapping$x)
  y <- GGally::eval_data_col(data, mapping$y)
  r <- cor(x, y, use = "complete.obs")
  ggplot2::ggplot() +
    ggplot2::annotate("text", x = 0.5, y = 0.5,
                      label = sprintf("%.2f", r),
                      size = 3.5) +
    ggplot2::xlim(0, 1) + ggplot2::ylim(0, 1) +
    theme_synth +
    ggplot2::theme(panel.grid.major = ggplot2::element_blank())
}
```

## Background

AI coding agents such as Claude Code, GitHub Copilot, and Cursor require
access to data to develop analysis pipelines. When the data contain
patient-identifiable records, this poses a problem: the agent transmits
data to a remote API. Deny rules in the agent's configuration can block
direct file reads, but shell output, error messages, and script results
may still expose real values in the conversation context.

One approach is to replace the real data with a synthetic version that
preserves the schema and statistical properties of the original. The
pipeline is developed against the synthetic data; the final analysis is
run on the real data outside the agent.

This vignette demonstrates the relevant `synthpop` functions: synthesis,
diagnostic comparison, disclosure control, and project integration.

## Synthesising a dataset

`syn()` fits a sequence of conditional models to the real data (one per
column, in visit-sequence order) and draws new records from those models.

```{r}
#| warning: true
#| message: true
library(synthpop)

syn_mtcars <- syn(mtcars, method = "parametric", seed = 2024)
```

`synthpop` warns that `mtcars` has fewer observations (32) than the
recommended minimum of 210 (100 + 10 per variable). The sequential
conditional models used by `syn()` require enough observations to
estimate each column's relationship with its predecessors. With too few
rows relative to columns, the models are unstable and the synthetic data
may not reproduce the original distributions or correlations reliably.
Wide datasets (many columns, few rows) are particularly problematic
because later variables in the visit sequence are conditioned on all
preceding columns, compounding the estimation difficulty. The diagnostic
functions shown below are important for detecting these failures.

The result is a `synds` object. The synthetic data frame is stored in `$syn`:

```{r}
identical(dim(syn_mtcars$syn), dim(mtcars))
identical(names(syn_mtcars$syn), names(mtcars))
```

### Preparing categorical variables

`synthpop` warns when numeric columns have 5 or fewer unique values,
since these are typically categorical. Converting them to factors before
synthesis ensures they are modelled with logistic or multinomial models
rather than linear ones, and that synthetic values remain within the
observed levels.

`numtocat.syn()` does this automatically. Manual conversion is also
possible:

```{r}
mtcars_prep <- mtcars
mtcars_prep$cyl <- factor(mtcars_prep$cyl)
mtcars_prep$vs <- factor(mtcars_prep$vs)
mtcars_prep$am <- factor(mtcars_prep$am)
mtcars_prep$gear <- factor(mtcars_prep$gear)

syn_mtcars_f <- syn(mtcars_prep, method = "parametric", seed = 2024)
```

The synthetic data retains the correct factor levels:

```{r}
table(syn_mtcars_f$syn$cyl)
table(mtcars_prep$cyl)
```

## Choosing a method

The `method` argument selects the synthesis model. `"parametric"` fits
linear models for numeric columns and logistic regression for factors;
it preserves pairwise correlations well and is faster. `"cart"` (the
default) fits classification and regression trees, which can capture
non-linear relationships and subgroup structure. Methods can be set
per column.

```{r}
syn_param <- syn(mtcars_prep, method = "parametric", seed = 2024)
syn_cart <- syn(mtcars_prep, method = "cart", seed = 2024)
```

## Comparing synthetic and real data

### Pairwise comparison

`GGally::ggpairs` overlays real and synthetic data in a pairs matrix.
The upper triangle contains correlations, the diagonal contains density
overlays, and the lower triangle contains scatterplots.

```{r}
#| label: fig-pairwise-mtcars
#| fig-cap: "Pairwise comparison of real and parametric synthetic mtcars data."
#| fig-alt: "Pairs matrix of five numeric variables (mpg, cyl, disp, hp, wt). Upper triangle shows correlation coefficients, diagonal shows density overlays, and lower triangle shows scatter plots. Real data in blue, synthetic in orange. Density shapes and correlations are closely matched between real and synthetic."
#| fig-width: 12
#| fig-height: 10
if (requireNamespace("GGally", quietly = TRUE)) {
  num_plot_cols <- c("mpg", "cyl", "disp", "hp", "wt")
  real_df <- mtcars[, num_plot_cols]
  syn_df <- syn_param$syn[, num_plot_cols]
  syn_df$cyl <- as.numeric(as.character(syn_df$cyl))
  real_df$.source <- "Real"
  syn_df$.source <- "Synthetic"
  combined <- rbind(real_df, syn_df)
  combined$.source <- factor(combined$.source,
                             levels = c("Real", "Synthetic"))

  GGally::ggpairs(
    combined,
    columns = seq_along(num_plot_cols),
    mapping = ggplot2::aes(colour = .source, alpha = .source),
    upper = list(continuous = "cor"),
    lower = list(
      continuous = GGally::wrap("points", size = 1)
    ),
    diag = list(
      continuous = GGally::wrap("densityDiag", alpha = 0.4)
    ),
    title = "Real vs Synthetic: Pairwise Comparison",
    progress = FALSE,
    legend = c(3, 5)
  ) +
    ggplot2::scale_colour_manual(
      values = c(Real = col_real, Synthetic = col_synth)
    ) +
    ggplot2::scale_fill_manual(
      values = c(Real = col_real, Synthetic = col_synth)
    ) +
    ggplot2::scale_alpha_manual(
      values = c(Real = 0.6, Synthetic = 0.6), guide = "none"
    ) +
    theme_synth
}
```

For mixed-type data (numeric and factor columns), the pairs matrix
substitutes boxplots and bar charts where appropriate:

```{r}
#| label: fig-pairwise-iris
#| fig-cap: "Pairwise comparison of real and CART synthetic iris data, including the factor variable Species."
#| fig-alt: "Pairs matrix of four numeric iris measurements and Species. Scatter plots in the lower triangle show real and synthetic points overlaid with similar cluster separation. Density curves on the diagonal are closely matched. Boxplots show comparable distributions across species groups."
#| fig-width: 12
#| fig-height: 10
syn_iris <- syn(iris, method = "cart", seed = 2024)

if (requireNamespace("GGally", quietly = TRUE)) {
  real_ir <- iris
  syn_ir <- syn_iris$syn
  real_ir$Source <- "Real"
  syn_ir$Source <- "Synthetic"
  combined_ir <- rbind(real_ir, syn_ir)
  combined_ir$Source <- factor(combined_ir$Source, levels = c("Real", "Synthetic"))

  GGally::ggpairs(
    combined_ir,
    columns = 1:5,
    mapping = ggplot2::aes(colour = Source, alpha = Source),
    upper = list(continuous = cor_text,
                 combo = "blank", discrete = "blank"),
    lower = list(
      continuous = GGally::wrap("points", size = 1),
      combo = GGally::wrap("box_no_facet")
    ),
    diag = list(continuous = GGally::wrap("densityDiag", alpha = 0.4)),
    title = "Real vs Synthetic: iris (CART)",
    progress = FALSE,
    legend = c(1, 1)
  ) +
    ggplot2::scale_colour_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_fill_manual(values = c(Real = col_real, Synthetic = col_synth)) +
    ggplot2::scale_alpha_manual(values = c(Real = 0.6, Synthetic = 0.6), guide = "none") +
    theme_synth
}
```

### Per-variable comparison

`compare()` produces dodged bar charts for each variable, with the
$S_{pMSE}$ utility statistic printed in each panel title.

```{r}
#| label: fig-compare-mtcars
#| fig-cap: "Per-variable distribution comparison for parametric synthetic mtcars data."
#| fig-alt: "Faceted dodged bar chart with one panel per variable. Each panel shows observed and synthetic bars side by side. Most variables show close agreement between real and synthetic distributions."
#| fig-height: 10
compare(syn_param, mtcars_prep)
```

The same comparison for iris synthesised with CART:

```{r}
#| label: fig-compare-iris
#| fig-cap: "Per-variable distribution comparison for CART synthetic iris data."
#| fig-alt: "Faceted dodged bar chart comparing observed and synthetic iris data across five variables. Bar heights are closely matched. The Species panel shows near-identical proportions across the three levels."
#| fig-height: 8
compare(syn_iris, iris)
```

### Handling missingness

`synthpop` preserves missingness patterns. The `airquality` dataset has
`NA` values in `Ozone` and `Solar.R`:

```{r}
#| label: fig-compare-airquality
#| fig-cap: "Per-variable distribution comparison for parametric synthetic airquality data, including missing values."
#| fig-alt: "Faceted dodged bar chart for six airquality variables. Ozone and Solar.R panels include an NA category with similar proportions in real and synthetic data."
#| fig-height: 6
syn_aq <- syn(airquality, method = "parametric", seed = 2024)
compare(syn_aq, airquality)
```

```{r}
na_comparison <- data.frame(
  column = names(airquality),
  real_na = colSums(is.na(airquality)),
  synthetic_na = colSums(is.na(syn_aq$syn))
)
na_comparison[na_comparison$real_na > 0, ]
```

### Utility measures

`utility.gen()` trains a classifier to distinguish real from synthetic
records. `S_pMSE` is the propensity-score mean squared error standardised
by its null expectation; values near 1 indicate the classifier performs
no better than it would on a random split. `SPECKS` is the
Kolmogorov-Smirnov statistic on the propensity scores (0 =
indistinguishable, 0.5 = maximally distinguishable).

```{r}
ug_param <- utility.gen(syn_param, mtcars_prep, print.flag = FALSE)
ug_cart <- utility.gen(syn_cart, mtcars_prep, print.flag = FALSE)

data.frame(
  method = c("Parametric", "CART"),
  pMSE = round(c(ug_param$pMSE, ug_cart$pMSE), 4),
  S_pMSE = round(c(ug_param$S_pMSE, ug_cart$S_pMSE), 2),
  SPECKS = round(c(ug_param$SPECKS, ug_cart$SPECKS), 3)
)
```

### Correlation preservation

`synthpop` does not include a built-in correlation comparison. The mean
and maximum absolute differences in pairwise correlations can be computed
directly:

```{r}
num_cols <- names(mtcars_prep)[sapply(mtcars_prep, is.numeric)]
cor_real <- cor(mtcars_prep[num_cols], use = "complete.obs")
cor_param <- cor(syn_param$syn[num_cols], use = "complete.obs")
cor_cart <- cor(syn_cart$syn[num_cols], use = "complete.obs")

idx <- lower.tri(cor_real)

data.frame(
  method = c("Parametric", "CART"),
  mean_abs_diff = round(c(
    mean(abs(cor_real[idx] - cor_param[idx])),
    mean(abs(cor_real[idx] - cor_cart[idx]))
  ), 3),
  max_abs_diff = round(c(
    max(abs(cor_real[idx] - cor_param[idx])),
    max(abs(cor_real[idx] - cor_cart[idx]))
  ), 3)
)
```

### Comparing model fits

`lm.synds()` fits a linear model to the synthetic data. `compare()` then
plots the coefficients alongside the real-data fit. Overlapping confidence
intervals indicate the synthetic data preserves the modelled relationships.

```{r}
#| label: fig-compare-fit
#| fig-cap: "Coefficient comparison for a linear model (mpg ~ wt + hp + cyl) fitted to real and synthetic data."
#| fig-alt: "Horizontal dot-and-whisker plot comparing regression coefficients from observed and synthetic linear models. Each coefficient has a point estimate and 95% confidence interval. The intervals overlap for all coefficients."
#| fig-width: 8
#| fig-height: 5
fit_syn <- lm.synds(mpg ~ wt + hp + cyl, data = syn_param)
compare(fit_syn, mtcars_prep)
```

## Disclosure control

### Replicated uniques

A replicated unique is a synthetic record that matches a unique real
record on all columns. These represent the most direct re-identification
risk.

```{r}
ru_param <- replicated.uniques(syn_param, mtcars_prep)
ru_cart <- replicated.uniques(syn_cart, mtcars_prep)

data.frame(
  method = c("Parametric", "CART"),
  replicated_uniques = c(
    ru_param$res_tab["Replicated uniques", "Number"],
    ru_cart$res_tab["Replicated uniques", "Number"]
  ),
  total_records = nrow(mtcars_prep)
)
```

CART tends to produce more replicated uniques because tree leaves can
reproduce small groups of records. This is more pronounced in datasets
with well-separated clusters:

```{r}
ru_iris <- replicated.uniques(syn_iris, iris)

data.frame(
  dataset = "iris (CART)",
  replicated_uniques = ru_iris$res_tab["Replicated uniques", "Number"],
  total_records = nrow(iris),
  pct = round(ru_iris$res_tab["Replicated uniques", "%"], 1)
)
```

### Identity and attribute disclosure

`disclosure()` quantifies whether knowing a set of quasi-identifier
columns (keys) allows an attacker to identify an individual or infer a
sensitive attribute:

```{r}
disc <- disclosure(syn_param, mtcars_prep,
                   keys = c("cyl", "hp", "wt"), target = "mpg")
```

```{r}
data.frame(
  measure = c("Identity disclosure (%)", "Attribute disclosure (%)"),
  baseline = round(c(disc$ident[1, 1], disc$attrib[1, 1]), 1),
  synthetic = round(c(disc$ident[2, 1], disc$attrib[2, 1]), 1)
)
```

Synthetic disclosure rates well below the baseline indicate low risk.

### Removing high-risk records

`sdc()` removes or replaces replicated uniques:

```{r}
syn_iris_sdc <- sdc(syn_iris, iris, rm.replicated.uniques = TRUE)
ru_after <- replicated.uniques(syn_iris_sdc, iris)

data.frame(
  stage = c("Before SDC", "After SDC"),
  replicated_uniques = c(
    ru_iris$res_tab["Replicated uniques", "Number"],
    ru_after$res_tab["Replicated uniques", "Number"]
  )
)
```

## Longitudinal and panel data

`synthpop` operates on a single data frame: one row per record, one
column per variable. Longitudinal data, where each individual
contributes multiple timepoints, does not fit this layout directly.
Three strategies are available, each with different trade-offs between
fidelity, speed, and scalability.

### Simulated cohort

The examples below use a simulated cohort of 500 individuals observed
over 30 days, with a viral load trajectory that depends on age and
vaccination status.

```{r}
#| label: longitudinal-data
set.seed(42)

n_individuals <- 500
n_timepoints <- 30

pop <- data.frame(
  id = 1:n_individuals,
  age = round(rnorm(n_individuals, mean = 45, sd = 15)),
  sex = factor(sample(c("M", "F"), n_individuals, replace = TRUE)),
  vaccinated = factor(sample(c(0, 1), n_individuals, replace = TRUE,
                             prob = c(0.4, 0.6)))
)

generate_vl <- function(age, vaccinated) {
  peak_day <- rnorm(1, mean = 5, sd = 1.2)
  peak_vl <- rnorm(1, mean = 7.5 - 0.8 * vaccinated - 0.01 * age, sd = 0.5)
  decay_rate <- rnorm(1, mean = 0.3 + 0.1 * vaccinated, sd = 0.05)
  growth_rate <- rnorm(1, mean = 1.2, sd = 0.2)
  days <- 1:n_timepoints
  vl <- peak_vl - growth_rate * pmax(0, peak_day - days) -
    decay_rate * pmax(0, days - peak_day)
  pmax(0, vl + rnorm(n_timepoints, 0, 0.3))
}

vl_list <- mapply(generate_vl, pop$age, as.integer(as.character(pop$vaccinated)),
                  SIMPLIFY = FALSE)
vl_long <- do.call(rbind, lapply(seq_along(vl_list), function(i) {
  data.frame(id = i, day = 1:n_timepoints, vl = vl_list[[i]])
}))
vl_long <- merge(vl_long, pop, by = "id")
```

Two derived objects are used throughout: a wide version (one row per
individual, one column per day) and a helper function that computes the
mean within-individual lag-1 autocorrelation, which measures how well
the synthetic data preserves temporal dependence.

```{r}
#| label: longitudinal-helpers
vl_wide <- reshape(
  vl_long[, c("id", "day", "vl", "age", "sex", "vaccinated")],
  direction = "wide", idvar = c("id", "age", "sex", "vaccinated"),
  timevar = "day", v.names = "vl"
)
names(vl_wide) <- sub("^vl\\.", "vl_day", names(vl_wide))

lag1_autocor <- function(long_df) {
  ids <- unique(long_df$id)
  cors <- vapply(ids, function(i) {
    x <- long_df$vl[long_df$id == i]
    x <- x[order(long_df$day[long_df$id == i])]
    cor(x[-length(x)], x[-1], use = "complete.obs")
  }, numeric(1))
  mean(cors, na.rm = TRUE)
}

real_lag1 <- lag1_autocor(vl_long)
```

### Strategy 1: Wide format

Pivoting to wide format so that each individual occupies one row is the
most direct approach. Each timepoint becomes a column, and the
sequential conditional models in `syn()` capture the temporal
correlations as cross-column dependencies.

```{r}
#| label: longitudinal-wide
wide_df <- vl_wide[, setdiff(names(vl_wide), "id")]

t_wide <- system.time({
  syn_wide <- syn(wide_df, method = "cart", seed = 2024)
})

syn_wide_long <- reshape(
  syn_wide$syn, direction = "long",
  varying = grep("^vl_day", names(syn_wide$syn), value = TRUE),
  v.names = "vl", timevar = "day", idvar = "syn_id"
)
syn_wide_long$id <- syn_wide_long$syn_id

cat("Time:", round(t_wide["elapsed"], 1), "s\n")
cat("Lag-1 autocorrelation - real:", round(real_lag1, 3),
    " synthetic:", round(lag1_autocor(syn_wide_long), 3), "\n")
```

This preserves temporal structure well because adjacent columns are
highly correlated and CART splits on them effectively. The limitation is
column count: with $T$ timepoints and $p$ baseline covariates, the data
frame has $p + T$ columns. `syn()` conditions each column on all
preceding ones, so the model complexity grows quadratically with $T$.
For moderate $T$ (up to a few dozen timepoints), this works. For
hundreds of timepoints, the synthesis becomes slow or unstable.

### Strategy 2: Wide format with parametric method

Replacing `method = "cart"` with `method = "parametric"` fits linear
regression (for numeric columns) rather than trees. Linear models handle
many predictors more gracefully and run faster, at the cost of assuming
linear relationships between columns.

```{r}
#| label: longitudinal-parametric
t_param <- system.time({
  syn_param_wide <- syn(wide_df, method = "parametric", seed = 2024)
})

syn_pw_long <- reshape(
  syn_param_wide$syn, direction = "long",
  varying = grep("^vl_day", names(syn_param_wide$syn), value = TRUE),
  v.names = "vl", timevar = "day", idvar = "syn_id"
)
syn_pw_long$id <- syn_pw_long$syn_id

cat("Time:", round(t_param["elapsed"], 1), "s\n")
cat("Lag-1 autocorrelation - real:", round(real_lag1, 3),
    " synthetic:", round(lag1_autocor(syn_pw_long), 3), "\n")
```

For longitudinal data with smooth trajectories, parametric synthesis
performs comparably to CART at a fraction of the time. It is the
recommended starting point when the number of timepoints is large
relative to the number of individuals.

### Strategy 3: Two-stage synthesis

When the wide format is too large, an alternative is to summarise each
individual's trajectory into a small number of parameters (peak value,
peak timing, area under the curve), synthesise those summaries alongside
the baseline covariates, and then reconstruct timepoints from the
synthetic summaries.

```{r}
#| label: longitudinal-twostage
traj_summaries <- do.call(rbind, lapply(split(vl_long, vl_long$id), function(d) {
  data.frame(
    id = d$id[1],
    vl_peak = max(d$vl),
    vl_peak_day = d$day[which.max(d$vl)],
    vl_mean = mean(d$vl),
    vl_last = d$vl[nrow(d)]
  )
}))

baseline_summ <- merge(pop[, setdiff(names(pop), "id"), drop = FALSE],
                       traj_summaries[, setdiff(names(traj_summaries), "id")],
                       by = "row.names")[, -1]

t_twostage <- system.time({
  syn_summ <- syn(baseline_summ, method = "cart", seed = 2024)
})

reconstruct_trajectory <- function(peak_vl, peak_day, n_days) {
  peak_day <- max(1, min(peak_day, n_days))
  growth <- peak_vl / peak_day
  decay <- peak_vl / (n_days - peak_day + 1)
  days <- 1:n_days
  vl <- ifelse(days <= peak_day,
               growth * days,
               peak_vl - decay * (days - peak_day))
  pmax(0, vl + rnorm(n_days, 0, 0.3))
}

set.seed(99)
syn_traj_long <- do.call(rbind, lapply(seq_len(nrow(syn_summ$syn)), function(i) {
  row <- syn_summ$syn[i, ]
  data.frame(
    id = i, day = 1:n_timepoints,
    vl = reconstruct_trajectory(row$vl_peak, row$vl_peak_day, n_timepoints)
  )
}))

cat("Time:", round(t_twostage["elapsed"], 1), "s\n")
cat("Lag-1 autocorrelation - real:", round(real_lag1, 3),
    " synthetic:", round(lag1_autocor(syn_traj_long), 3), "\n")
```

The two-stage approach is fast and scales to any number of timepoints,
because the data frame passed to `syn()` has only a handful of columns
regardless of $T$. The trade-off is that the reconstructed trajectories
are simplified: the parametric reconstruction imposes a shape (here, a
linear rise and decay) that may not match the real trajectories. For the
purpose of developing an analysis pipeline against synthetic data, this
is often acceptable. The reconstruction function should be chosen to
match the expected shape of the real data.

### Comparison

```{r}
#| label: fig-longitudinal-comparison
#| fig-cap: "Within-individual lag-1 autocorrelation and runtime for three longitudinal synthesis strategies."
#| fig-alt: "Bar chart comparing three strategies (wide CART, wide parametric, two-stage) on lag-1 autocorrelation and runtime. Wide formats preserve autocorrelation near the real value of 0.95. Two-stage is slightly lower at 0.93 but faster."
#| fig-width: 10
#| fig-height: 5
strategies <- data.frame(
  strategy = c("Real data", "Wide (CART)", "Wide (parametric)", "Two-stage"),
  lag1 = c(
    real_lag1,
    lag1_autocor(syn_wide_long),
    lag1_autocor(syn_pw_long),
    lag1_autocor(syn_traj_long)
  ),
  time = c(0, t_wide["elapsed"], t_param["elapsed"], t_twostage["elapsed"])
)

par(mfrow = c(1, 2), mar = c(5, 8, 3, 1))

barplot(
  strategies$lag1, names.arg = strategies$strategy,
  horiz = TRUE, las = 1, col = c("grey70", col_real, col_real, col_synth),
  xlab = "Lag-1 autocorrelation",
  main = "Temporal fidelity", xlim = c(0, 1)
)
abline(v = real_lag1, lty = 2, col = "grey40")

barplot(
  strategies$time[-1], names.arg = strategies$strategy[-1],
  horiz = TRUE, las = 1, col = c(col_real, col_real, col_synth),
  xlab = "Seconds",
  main = "Runtime"
)
```

### Individual trajectories

Plotting a sample of individual trajectories from the real and synthetic
data (using the wide-format CART synthesis) shows whether the synthetic
records reproduce realistic trajectory shapes.

```{r}
#| label: fig-longitudinal-trajectories
#| fig-cap: "Sample of 20 individual viral load trajectories from the real and wide-format CART synthetic data."
#| fig-alt: "Line plot showing 20 real trajectories in blue and 20 synthetic trajectories in orange. Both sets show a rise to a peak around day 5 followed by a gradual decline, with similar spread and shape."
#| fig-width: 10
#| fig-height: 6
set.seed(1)
sample_real <- vl_long[vl_long$id %in% sample(unique(vl_long$id), 20), ]
sample_real$source <- "Real"

sample_syn <- syn_wide_long[
  syn_wide_long$id %in% sample(unique(syn_wide_long$id), 20),
]
sample_syn$source <- "Synthetic"

plot_cols <- c("id", "day", "vl", "source")
plot_data <- rbind(sample_real[, plot_cols], sample_syn[, plot_cols])

plot(NULL, xlim = c(1, n_timepoints), ylim = range(plot_data$vl),
     xlab = "Day", ylab = "Viral load (log10 copies/mL)",
     main = "Individual trajectories: real vs synthetic (wide CART)")

for (src in c("Real", "Synthetic")) {
  col <- if (src == "Real") col_real else col_synth
  sub <- plot_data[plot_data$source == src, ]
  for (i in unique(sub$id)) {
    d <- sub[sub$id == i, ]
    d <- d[order(d$day), ]
    lines(d$day, d$vl, col = adjustcolor(col, alpha.f = 0.5))
  }
}
legend("topright", legend = c("Real", "Synthetic"),
       col = c(col_real, col_synth), lwd = 2, bty = "n")
```

### Mean trajectory by vaccination status

Averaging across individuals within each vaccination group reveals
whether the synthesis preserves the group-level treatment effect.

```{r}
#| label: fig-longitudinal-mean-vax
#| fig-cap: "Mean viral load trajectory by vaccination status for real and wide-format CART synthetic data."
#| fig-alt: "Line plot with four lines: real vaccinated, real unvaccinated, synthetic vaccinated, and synthetic unvaccinated. Unvaccinated groups peak higher than vaccinated groups. Real and synthetic lines are closely matched within each group."
#| fig-width: 10
#| fig-height: 6
mean_by_vax <- function(long_df, vacc_col = "vaccinated") {
  vacc_vals <- unique(long_df[[vacc_col]])
  do.call(rbind, lapply(vacc_vals, function(v) {
    sub <- long_df[long_df[[vacc_col]] == v, ]
    agg <- aggregate(vl ~ day, data = sub, FUN = mean)
    agg$vaccinated <- v
    agg
  }))
}

mean_real <- mean_by_vax(vl_long)
mean_real$source <- "Real"

syn_wide_long$vaccinated <- rep(
  as.character(syn_wide$syn$vaccinated),
  each = n_timepoints
)
mean_syn <- mean_by_vax(syn_wide_long)
mean_syn$source <- "Synthetic"

mean_data <- rbind(mean_real, mean_syn)

plot(NULL, xlim = c(1, n_timepoints), ylim = range(mean_data$vl),
     xlab = "Day", ylab = "Mean viral load (log10 copies/mL)",
     main = "Mean trajectory by vaccination status")

vacc_levels <- unique(mean_data$vaccinated)
lty_map <- setNames(seq_along(vacc_levels), vacc_levels)

for (src in c("Real", "Synthetic")) {
  col <- if (src == "Real") col_real else col_synth
  for (v in vacc_levels) {
    sub <- mean_data[mean_data$source == src &
                       mean_data$vaccinated == v, ]
    sub <- sub[order(sub$day), ]
    lines(sub$day, sub$vl, col = col, lwd = 2,
          lty = lty_map[as.character(v)])
  }
}
legend("topright",
       legend = c("Real - unvacc", "Real - vacc",
                  "Synthetic - unvacc", "Synthetic - vacc"),
       col = rep(c(col_real, col_synth), each = 2),
       lty = rep(1:2, 2), lwd = 2, bty = "n")
```

### What does not work

Two approaches that appear natural tend to fail in practice.

**Synthesising in long format** (one row per individual-timepoint)
destroys within-individual correlation. `syn()` treats each row as
independent, so the temporal ordering within an individual is lost. The
lag-1 autocorrelation drops sharply:

```{r}
#| label: longitudinal-long-naive
long_df <- vl_long[, c("age", "sex", "vaccinated", "day", "vl")]
syn_long_naive <- syn(long_df, method = "cart", seed = 2024)

syn_long_naive_df <- syn_long_naive$syn
syn_long_naive_df$id <- rep(1:(nrow(syn_long_naive_df) %/% n_timepoints),
                            each = n_timepoints)[seq_len(nrow(syn_long_naive_df))]

cat("Lag-1 autocorrelation - real:", round(real_lag1, 3),
    " long-format synthetic:", round(lag1_autocor(syn_long_naive_df), 3), "\n")
```

**Looping over cluster units** (synthesising each cluster or individual
separately in wide format) splits the data into subsets that are often
too small for the number of columns. `syn()` warns when $n < 100 + 10p$,
and with many timepoint columns each subset will fall well below that
threshold. The synthesis is unstable, slow in aggregate, and may exhaust
memory for large cohorts because each call builds its own tree ensemble.

### Recommendations

For longitudinal data with up to a few dozen timepoints, pivot to wide
format and use `method = "parametric"`. It is fast, stable, and
preserves within-individual temporal structure well.

For data with many timepoints or very wide covariate sets, use the
two-stage approach: synthesise trajectory summaries alongside baseline
covariates, then reconstruct the time series. Choose summary parameters
and a reconstruction function that match the expected shape of the
real trajectories.

Do not synthesise in long format (it destroys temporal dependence) or
loop over individuals or small clusters (it is slow and unstable).

## Data governance considerations

Synthetic data is not anonymised data. The following should be addressed
before sending synthetic data to an external API.

**Individual-level risk.** Direct identifiers (names, addresses, NHS
numbers, dates of birth) must be removed before synthesis.
`replicated.uniques()` and `disclosure()` should be run on the output,
and `sdc()` used to remove any replicated uniques.

**Population-level risk.** Synthetic data preserves the joint
distributions and correlations of the original cohort. For large
datasets on common conditions, this is unlikely to be sensitive. For
small or rare-disease cohorts at a named institution, the statistical
structure itself may be sensitive: a synthetic dataset that preserves the
relationship between genotype and outcome for 12 patients carries the
same information as the real data for anyone who knows the cohort exists.
Synthesis does not resolve the governance problem in such cases. As a
rule of thumb, if you would be comfortable sharing detailed summary
statistics and cross-tabulations with the API, the synthetic data is
comparable in information content.

**What your DPO needs to see.** A description of the synthesis method,
confirmation that identifiers were removed, the `replicated.uniques()`
and `disclosure()` output, the data processor's retention and
training-use policies, and a documented risk assessment.

## Integration into an analysis workflow

### Generation script

Run this outside the agent, in a regular R session with access to the
real data:

```r
# scripts/00_generate_synthetic.R
library(here)
library(synthpop)
library(readr)

raw_dir <- here("data", "raw")
syn_dir <- here("data", "synthetic")
if (!dir.exists(syn_dir)) dir.create(syn_dir, recursive = TRUE)

for (file in list.files(raw_dir, pattern = "\\.csv$", full.names = TRUE)) {
  real <- read_csv(file, show_col_types = FALSE)
  result <- syn(as.data.frame(real), method = "parametric", seed = 2024)

  ru <- replicated.uniques(result, as.data.frame(real))
  if (ru$res_tab["Replicated uniques", "Number"] > 0) {
    result <- sdc(result, as.data.frame(real), rm.replicated.uniques = TRUE)
  }

  write_csv(result$syn, file.path(syn_dir, basename(file)))
}
```

### Switching between synthetic and real data

An environment variable controls which data directory is used, with the
safe default being synthetic:

```r
use_synthetic <- toupper(Sys.getenv("SYNTHETIC_DATA", "TRUE")) %in%
  c("TRUE", "1", "YES", "")

data_dir <- if (use_synthetic) {
  here::here("data", "synthetic")
} else {
  here::here("data", "raw")
}
```

To run on real data:

```bash
SYNTHETIC_DATA=false Rscript scripts/00_run_pipeline.R
```

## Summary

| synthpop function | Purpose |
|---|---|
| `syn()` | Generate synthetic data |
| `compare()` | Per-variable distribution comparison |
| `lm.synds()` / `glm.synds()` | Fit models to synthetic data |
| `compare.fit.synds()` | Compare model fits (real vs synthetic) |
| `utility.gen()` | Propensity score utility measure |
| `replicated.uniques()` | Check for records matching unique real individuals |
| `disclosure()` | Identity and attribute disclosure risk |
| `sdc()` | Remove or replace high-risk synthetic records |
